Forecasting the Support of Kamala Harris in 2024 U.S. Presidential Election using a Model-Based Approach

### 2024-11-04 paper/ForecastingtheSupportofKamalaHarrisin2024U.S.PresidentialElection.qmd
update paper

The linear regression models constructed for predicting support for
Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election
utilized both unweighted and weighted configurations. The model summary of support for Harris is shown in @tbl-summary1, and support for Trump is shown in @tbl-summary2. As shown in
@tbl-summary1 and @tbl-summary2, the intercepts for Harris and Trump,
estimated at 44.70% and 51.87% respectively in the unweighted models, while in weighted models, it's estimated at 45.2% and 54.8%,
represent their baseline support levels when all predictors are at
reference values. The high ( $R^2$ ) values of 0.824 for Harris’s
unweighted model and 0.892 for her weighted model, along with a similar
values of 0.849 and 0.916 for Trump, indicate that a significant portion
of the variation in support can be explained by the model's predictors.
The relatively low Root Mean Square Error (RMSE) values further suggest
that the models provide reasonably accurate predictions of support
levels.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average of 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)

##weighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted), size = 0.2)+
  labs(x = "End Date of the Poll",
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))  +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| fig-cap: "Harris is likely to lead Trump in Pennsylvania, Michigan, and Wisconsin by Model Prediction (From August 1, 2024, to November 2, 2024)."
#| label: fig-support-state
## unweighted
plan(multisession)
selected_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")
combined_data_total<-combined_data_total|>filter(state %in% selected_states)
combined_predictions_filtered<-combined_predictions_filtered|>filter(state %in% selected_states)
late<-combined_predictions_filtered[which.max(combined_predictions_filtered$end_date), ]
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate))+
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),se = FALSE, size = 0.5)+
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, size =6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

## weighetd
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted),se = FALSE, size = 0.5)+ # Smoothed trend line
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1,size=6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) 
```

@@

ggplot(combined_data_total, aes(x = end_date, y = pct)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_unweighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_unweighted), color = "red", size = 1) +
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    color = "Candidate",
    fill = "Top 5 Pollster",
    caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) + 
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

ggplot(combined_data_total, aes(x = end_date, y = pct, size = combined_weight)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_weighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_weighted), color = "red", size = 1)+
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    size = "Weight",
    fill = "Top 5 Pollster",
    caption = "Lines are weighted by recency_weight * sample_size_weight.\nPoints are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatetst Poll: Nov 2\nSource: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

>>>>

The linear regression models constructed for predicting support for
Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election
utilized both unweighted and weighted configurations. The model summary of support for Harris is shown in @tbl-summary1, and support for Trump is shown in @tbl-summary2. As shown in
@tbl-summary1 and @tbl-summary2, the intercepts for Harris and Trump,
estimated at 44.70% and 51.87% respectively in the unweighted models, while in weighted models, it's estimated at 45.2% and 54.8%,
represent their baseline support levels when all predictors are at
reference values. The high $R^2$ values of 0.824 for Harris’s
unweighted model and 0.892 for her weighted model, along with a similar
values of 0.849 and 0.916 for Trump, indicate that a significant portion
of the variation in support can be explained by the model's predictors.
The relatively low Root Mean Square Error (RMSE) values further suggest
that the models provide reasonably accurate predictions of support
levels.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average of 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)

##weighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted), size = 0.2)+
  labs(x = "End Date of the Poll",
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight. Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red"))  +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| fig-cap: "Harris is likely to lead Trump in Pennsylvania, Michigan, and Wisconsin by Model Prediction (From August 1, 2024, to November 2, 2024)."
#| label: fig-support-state
## unweighted
plan(multisession)
selected_states <- c("Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin")
combined_data_total<-combined_data_total|>filter(state %in% selected_states)
combined_predictions_filtered<-combined_predictions_filtered|>filter(state %in% selected_states)
late<-combined_predictions_filtered[which.max(combined_predictions_filtered$end_date), ]
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate))+
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),se = FALSE, size = 0.5)+
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, size =6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

## weighetd
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate,size = combined_weight)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_weighted),se = FALSE, size = 0.5)+ # Smoothed trend line
  facet_wrap(~ state,ncol = 4) +
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",
       size="Weight",
       caption = "Lines are weighted by recency_weight * sample_size_weight. Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1,size=6),
        legend.box = "vertical",
        strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) 
```

@@

ggplot(combined_data_total, aes(x = end_date, y = pct)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_unweighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_unweighted), color = "red", size = 1) +
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    color = "Candidate",
    fill = "Top 5 Pollster",
    caption = "Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) + 
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))

ggplot(combined_data_total, aes(x = end_date, y = pct, size = combined_weight)) +
  geom_point(aes(fill = pollster_group), alpha = 0.7, shape = 21) + 
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Kamala Harris"), 
            aes(x = end_date, y = predicted_weighted), color = "blue", size = 1) +
  geom_smooth(data = combined_predictions_filtered %>% filter(candidate == "Donald Trump"), 
            aes(x = end_date, y = predicted_weighted), color = "red", size = 1)+
  labs(
    x = "End Date of the Poll",
    y = "Predicted Support (%)",
    size = "Weight",
    fill = "Top 5 Pollster",
    caption = "Lines are weighted by recency_weight * sample_size_weight. Points are polls.\nModel: Candidate Support Percentage = national_poll + pollster + population + state\nLatest Poll: Nov 2; Source: FiftyThirtyEight"
  )+ # Smoothed trend line
  facet_wrap(~ candidate,ncol = 2) +
  scale_fill_manual(values = c(setNames(rep("grey", length(unique(combined_data_total$pollster_group))), "Other"),
                               setNames(brewer.pal(10, "Set1"), top_pollsters))) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",  # Place legends in a horizontal row
    legend.title.align = 0.5,   # Center-align legend titles for better appearance
    axis.text.x = element_text(angle = 45, hjust = 1,size=8),
    strip.background = element_rect(colour = "black", fill = "white")
  ) +
  guides(
    size = guide_legend(order = 1, title.position = "top", label.position = "left"),
    fill = guide_legend(order = 2)
  ) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date))
